---
title: "Import & cleaning"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article focuses on the import from multiple files and participants, as well as the cleaning of the data. We need these packages:

```{r setup, message = FALSE}
library(LightLogR)
library(tidyverse)
library(gghighlight)
```

> Please note that this article uses the base pipe operator `|>`. You need an R version equal to or greater than 4.1.0 to use it. If you are using an older version, you can replace it with the `magrittr` pipe operator `%>%`.

# From which devices can I import data?

LightLogR aims to provide standard import routines for all devices used in research on personal light exposure. Currently, the following devices are supported:

```{r}
supported_devices()
```
More Information on these devices can be found in the reference for 
`import_Dataset()`.

## What if my device is not listed?

If you are using a device that is currently not supported, please contact the developers. We are always looking to expand the range of supported devices. The easiest and most trackable way to get in contact is by opening a new issue on our [Github repository](https://github.com/tscnlab/LightLogR/issues). Please see the [CONTRIBUTING](https://tscnlab.github.io/LightLogR/CONTRIBUTING.html) section for different ways to help and details about how this project handles them.

## What if my device is listed but the import does not work as expected?

We regularly find that files exported from the same device model can differ in structure. This may be due to different settings, software or hardware updates. If you encounter problems with the import, please get in contact with us, e.g. by opening an issue on our [Github repository](https://github.com/tscnlab/LightLogR/issues). Please see the [CONTRIBUTING](https://tscnlab.github.io/LightLogR/CONTRIBUTING.html) section for different ways to help and details about how this project handles them.

## Are there other ways to import data?

Yes. `LightLogR` simply requires a `data.frame` with a column containing `datetime` formatted data. Even a `light` data column is not strictly necessary, as `LightLogR` is optimized for, but not restricted to, light data. Further, an `Id` column is used in some functions to distinguish between different participants. 

To make life easier when using functions in `LightLogR`, the datetime column should be named `Datetime`, the id column `Id`, and, if present, the melanopic EDI light information `MEDI`.

Lastly, you can modify or add import functions that build upon `LightLogR`s import functionality. See the last chapter in this article for more information on that.

# Importing data

The first step in every analysis is data import. We will work with data collected as part of the Master Thesis *Insights into real-world human light exposure: relating self-report with eye-level light logging* by Carolina Guidolin (2023). The data is stored in 17 text files in the *data/* folder. You can access the data yourself through the [LightLogR GitHub repository](https://github.com/tscnlab/LightLogR/tree/main/vignettes/articles/data).

```{r, files}
#this assumes that you downloaded the files into a folder called "data" in the working directory
path <- "data"
files <- list.files(path, full.names = TRUE)
#show how many files are listes
length(files)
```

Next we require a time zone of data collection. If uncertain which time zones are valid, use the `OlsonNames()` function. Our data was collected in the "Europe/Berlin" time zone.

```{r, tz}
#first six time zones from OlsonNames()
head(OlsonNames())

#our time zone
tz <- "Europe/Berlin"
```

Lastly, the participant Ids are stored in the file names. We will extract them and store them in a column called `Id`. The following code defines the pattern as a *regular expression*, which will extract the first three digits from the file name.

```{r, Id pattern}
pattern <- "^(\\d{3})"
```

Now we can import the data. Data were collected with the ActLumus device by Condor Instruments. The right way to specify this is through the `import` function.

```{r, import}
data <- import$ActLumus(files, tz = tz, auto.id = pattern, print_n=33)
```

## My import is slow. Why is that and can I speed it up?

There are several possibilities, why the import is slow. The most common reasons are:

-   The data files are simply large. With short measurement intervals, many participants, and long measurement periods, files for a single study can easily reach several gigabytes. This takes some time to import and is ok.

-   The data files contain many gaps. During import, `LightLogR` checks for and visualizes gaps in the data. Especially large datasets with small intervals contain many gaps, which can slow down the import process.

-   The device model you are importing from has non-consistent data structures. Some devices have a varying number of rows before the actual data starts. This means a small portion of every file has to be read in and the correct starting row has to be found. This can slow down the import process if you have many files.

If you are experiencing slow imports, you can try the following:

-   Only read in part of your datasets, or split your dataset into several pieces, that each gets loaded in separately. You can combine them afterwards with `join_datasets()`.

-   If you have many gaps in your data, you can set `auto.plot = FALSE` in the import function. This will eliminate the call to `gg_overview()`, which calculates and visualizes the gaps in the data.

# Finding and handling gaps and irregular data

Before we can dive into the analysis part, we need to make sure we have a clean dataset. The import summary shows us two problems with the data:

-   two files have data that crosses daylight saving time (DST) changes. Because the ActLumus device does not adjust for DST, we need to correct for this.

-   Multiple Ids have single datapoints at the beginning of the dataset with gaps before actual data collection starts. These are test measurements to check equipment, but must be removed from the dataset.

Let us first deal with the DST change. LightLogR has an in-built function to correct for this during import. We thus will re-import the data, but make the import silent as to not clutter the output.

```{r, dst change}
data <- 
  import$ActLumus(files, tz = tz, auto.id = pattern, dst_adjustment = TRUE, silent = TRUE)
```

The second problem requires the filtering of certain Ids. The `filter_Datetime_multiple()` function is ideal for this. We can provide a length (1 week), starting from the end of data collection and backwards. The variable `arguments` provide variable arguments to the filter function, they have to be provided in list form and expressions have to be quoted through`quote()`. Fixed arguments, like the length and`length_from_start\` are provided as named arguments and only have to be specified once, as they are the same for all Ids.

```{r, start shift}
data <- 
  data |> 
  filter_Datetime_multiple(
    arguments = list(
      list(only_Id = quote(Id == 216)),
      list(only_Id = quote(Id == 219)),
      list(only_Id = quote(Id == 214)),
      list(only_Id = quote(Id == 206))
    ), 
    length = "1 week", length_from_start = FALSE)

```

Let's have a look at the data again with the `gg_overview()` function.

```{r, overview}
data |> gg_overview()
```

Looks much better now. Also, because there is no longer a hint about gaps in the lower right corner, we can be sure that all gaps have been removed. The function `has_irregulars()` shows us, however, that there are still irregularities in the data and the function `gap_table()` reveals where they are.

```{r, irregularities}
data |> 
  has_irregulars()
```

Having irregular data in a dataset can be very problematic, as it messes up many functions that require constant spacing of observations. Also, it can seriously slow down computational performance. Let us first check out where the irregular data happen, with the function `gap_table()`. It is important to set `include.implicit.gaps = FALSE`, or it will be computationally very costly

```{r, irregularities2}
data |> 
  gap_table(include.implicit.gaps = FALSE)
```

This means we have to look at and take care of the irregularities for the Ids 215, 218, and 221. Let us first visualize where the irregularities are. We can use `gg_gaps()` for that. We will use `include.implicit.gaps = FALSE` again to save computational cost, and also `group.by.days = TRUE`, wich will give us results by day.

```{r}
data |> gg_gaps(include.implicit.gaps = FALSE, group.by.days = TRUE, show.irregulars = TRUE)
```

The first instances seem fine, but then right after the start all data appear shifted. As the first half of the day is missing anyways, this is likely not a day we will be using. This is also the case with all the other participants, anyways, which we can show by selecting a sample and extract the gaps they have. 

```{r}
data |> 
  filter(Id %in% 208:210) |> 
  gg_gaps(MEDI, x.axis.format = "%a")

data |> 
  filter(Id %in% 208:210) |> 
  extract_gaps()

```

We can remove these partial days via `remove_partial_data()`. We want to make sure to apply this filter `by.date`, as it otherwise would outright remove participants, but we want to simply remove single days. Before we remove these days we can check which days are going to be removed.

```{r}

data |> 
  remove_partial_data(
    handle.gaps = TRUE, MEDI, by.date = TRUE, threshold.missing = "2 hours",
    show.result = TRUE
    ) |> 
  filter(marked.for.removal)

data <- 
  data |> 
    remove_partial_data(
      handle.gaps = TRUE, MEDI, by.date = TRUE, threshold.missing = "2 hours"
      )
```

34 days were removed. As we have 17 participants, that equals the first and last day of data collection, which are correctly identified as only partly available.

```{r}
data |> has_gaps()
data |> has_irregulars()
```

The data is now clean and we can proceed with the analysis. This dataset will be needed in other articles, so we will save it as an RDS file.

```{r}
# uncomment next lines to save the data
# if (!dir.exists("cleaned_data")) dir.create("cleaned_data")
# saveRDS(data, "cleaned_data/ll_data.rds")
```

# Importing data: Miscellaneous

## Other import arguments

Other potentially important arguments are the `locale` argument - this is useful if you have special characters in your data (e.g. German ü or ä) that are not recognized by the default locale. Look at `readr::default_locale()` for more information.

The `...` argument is passed through to whichever import function is used for the data. For some devices, it is also used to provide additional information, such as `column_names` for the Actiwatch devices, that differ depending on the language setting of the device and software. Whether a device requires additional information can be found in the import documentation (see `import_Dataset()`).

## Other ways to call import

Instead of using the import function as described above (`import$device()`), you can also use the function `import_Dataset()` and specify the device as a character string in the first argument. This might be useful if you want to import data programmatically from different devices, e.g., through a `purrr::map()` function. Only `supported_devices()` will be accepted by the function.

Here is an example:

```
devices <- c("ActLumus", "Speccy")
files_AL <- c("path/to/ActLumus/file1.csv", "path/to/ActLumus/file2.csv")
files_Sy <- c("path/to/Speccy/file1.csv", "path/to/Speccy/file2.csv")
tz <- "Europe/Berlin"

data <- purrr::map2(devices, list(files_AL, files_Sy), import_Dataset, tz = tz)
```

This way, you will end with a list of two dataframes, one with the ActLumus data and one with the Speccy data.

## Creating your own import function

**Note: This section is for advanced users only. You should be familiar with expressions in R and how to manipulate them.**

`LightLogR` comes with a number of custom import routines for different devices, that are then implemented into the main import function, which covers some general aspects and also creates the summary overview.

If you would like to write your own custom import function, `LightLogR` has you covered. First, you can see what makes all the import functions tick by looking at the included data set `ll_import_expr()`. This is a list of all the individual routines. Let's have a look at the ActLumus routine

```{r}
ll_import_expr()$ActLumus
```
We can see, it is rather simple, just a few lines of code. You can write your own expression and create an import function with it. The expression should create a `data` variable that contains the import script for the files. At the end of the expression, the `data` variable should contain the imported dataset and include a correctly formatted `Datetime` column, complete with the correct timezone.

Here we will create a variation of the old routine, that just adds a short message:

```{r}
new_import_expr <- ll_import_expr()
new_import_expr$ActLumus_new <- new_import_expr$ActLumus
new_import_expr$ActLumus_new[[6]] <- 
  rlang::expr({ cat("**Congratulation, you made a new import function**\n")
  data
  })
new_import_expr$ActLumus_new

```

We can now create a new import function with this expression. The function will be called `import$ActLumus_new()`.

```{r}
import <- import_adjustment(new_import_expr)
```

Let us now import a file of the previous dataset, setting the main summary and plotting function silent

```{r}
data <- import$ActLumus_new(files[1], tz = tz, auto.id = pattern, 
                         auto.plot = FALSE, silent = TRUE)
```

