---
title: "Visualizations"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article focuses on various ways to visualize personal light exposure data with `LightLogR`. It is important to note that `LightLogR` is using the `ggplot2` package for visualizations. This means that all the `ggplot2` functions can be used to further customize the plots.
The following packages are needed for the analysis:

```{r setup, message = FALSE}
library(LightLogR)
library(tidyverse)
library(patchwork)
```

> Please note that this article uses the base pipe operator `|>`. You need an R version equal to or greater than 4.1.0 to use it. If you are using an older version, you can replace it with the `magrittr` pipe operator `%>%`.

# Importing Data

We will use data imported and cleaned already in the article [Import & Cleaning](https://tscnlab.github.io/LightLogR/articles/Import.html).

```{r, import}
#this assumes the data is in the cleaned_data folder in the working directory
data <- readRDS("cleaned_data/ll_data.rds")
```

# gg_overview()

`gg_overview()` provides a glance at *when* data is available for each Id. Let's call it on our dataset.

```{r, overview}
data |> gg_overview()
```

As can be seen the dataset contains 17 ids with one weeks worth of data each, and one to three participants per week. `gg_overview()` will by default test whether there are gaps in the data and will show them as grey bars, as well as a message in the lower right corner. Let us force this behavior in our dataset by removing two days.

```{r}
data |>
  filter(
    !(date(Datetime) %in% c("2023-08-16", "2023-08-17"))
    ) |> 
  gg_overview()
```

Calculating gaps in the data can be computationally expensive for large datasets with small epochs. If you just require an overview of the data without being concerned about gaps, you can provide an empty `tibble::tibble()` to the `gap.data` argument. This will skip the gap calculation and speed up the graph generation.

```{r}
data |>
  filter(
    !(date(Datetime) %in% c("2023-08-16", "2023-08-17"))
    ) |> 
  gg_overview(gap.data = tibble())
```

*Hint: `gg_overview()` is automatically called by import functions in `LightLogR`, unless the argument `auto.plot = FALSE` is set. If your import is slow, this can also help in speeding up the process.*

# gg_day()

## Basics
`gg_day()` compares days within a dataset. By default it will use the `date`. Let's call it on a subset of our data. To distinguish between different Ids, we can set the `aes_col` argument to `Id`.

```{r, fig.width=7, fig.height = 10}
data |> 
  filter(Id %in% c(205, 206)) |> 
  gg_day(aes_col = Id, size = 0.5)
```

## Facetting

Note that each day is represented by its own facet, which is named after the date. We can give each Id its own facet by using the `ggplot2::facet_wrap()` function. The `Day.data` column is produced by `gg_day()` and contains the structure of the daily facets. It has to be used by `facet_wrap()` to ensure that the facets are shown correctly. We also reduce the breaks on the x-axis to avoid overlap at 00:00.

```{r, fig.width=7, fig.height = 10}
data |> 
  filter(Id %in% c(205, 206)) |> 
  gg_day(aes_col = Id, size = 0.5, 
         x.axis.breaks = hms::hms(hours = c(0, 6, 12, 18))) + 
  guides(color = "none") +
  facet_grid(rows = vars(Day.data), cols = vars(Id), switch = "y")
```

## Date-grouping

Showing the days by date is the default behavior of `gg_day()`. It can also be grouped by any other formatting of `base::strptime()`. Using `format.day = "%A"` in the function call will group all output by the weekday. Putting so many Participants in each facet makes the plot unreadable, but it demonstrates how `gg_day()` can be configured to combine observations from different dates. We have to provide a different color scale compared to the default one, as the default has only 10 colors compared to the 17 we need here.

```{r, fig.width=7, fig.height = 10}
data |> 
  gg_day(aes_col = Id, size = 0.5, format.day = "%A") + 
  scale_color_viridis_d()
```

## Customizing geoms and miscellanea

`gg_day()` uses `geom_point()` by default. This can be changed by providing a different `geom` to the function. Here we use `geom_line()` to connect the points. To make this more readable. Let us first recreate a simpler version of the above dataset by filtering and aggregating

```{r}
data_subset <- 
  data |> 
  filter(Id %in% c(205, 206)) |> #choosing 2 ids
  aggregate_Datetime(unit = "15 mins") |> #aggregating to 15 min intervals
  filter_Datetime(length = "3 days", full.day = TRUE) #restricting their length to 3 days

data_subset |> 
  gg_day(aes_col = Id)
```

Now we can use a different geom.
```{r}
data_subset |>
  gg_day(aes_col = Id, geom = "line")
```

Also a ribbon is possible.
```{r}
data_subset |>  
  gg_day(aes_col = Id, aes_fill = Id, geom = "ribbon", alpha = 0.5)
```

# gg_days()

This is the companion function to `gg_day()`. Instead of using individual days, it will create a timeline of days across all Ids.

```{r, fig.width=12}
data_subset2 <- 
data |> 
  filter(Id %in% c(205, 216, 219)) |> #choosing 3 ids
  aggregate_Datetime(unit = "15 mins") #aggregating to 15 min intervals

data_subset2 |>
  gg_days()
```

By default, `gg_days()` will always plot full days. Let us strip one participant of data for three days.

```{r, fig.width=12}
data_subset3 <- 
data_subset2 |> 
  filter(!(Id == 205 & 
             date(Datetime) %in% c("2023-08-29", "2023-08-30", "2023-08-31")))

data_subset3 |>
  gg_days()
```

You can see the plots are misaligned in their facets. We can correct for that by providing an exact number of days to the `x.axis.limits` argument. `Datetime_limits()` is a helper function from `LightLogR` and the documentation reveals more about its arguments.

```{r, fig.width=12}
data_subset3 |> 
  gg_days(
    x.axis.limits = 
      \(x) Datetime_limits(x, length = ddays(5), midnight.rollover = TRUE)
    )
```

`gg_days()` has all of the customization options of `gg_day()`. Here we will customize the plot for a ribbon, different naming and breaks on the datetime axis.

```{r, fig.width=10}
data_subset3 |> 
  gg_days(
    geom = "ribbon", aes_col = Id, aes_fill = Id, alpha = 0.5, jco_color = TRUE,
    x.axis.limits = 
      \(x) Datetime_limits(x, length = ddays(5), midnight.rollover = TRUE),
    x.axis.breaks = 
      \(x) Datetime_breaks(x, by = "6 hours", shift = 0),
    x.axis.format = "%H",
    ) +
  guides(color = "none", fill = "none")
```

# gg_heatmap()

`gg_heatmap()` is a simple but handy feature to get an overview over large time periods across groups. (For some unknown reason, it tends to produce a warning about uneven horizontal intervals, which can be safely ignored)

```{r, fig.height = 12, fig.width = 20, warning = FALSE}
data |> gg_heatmap()
```

The function does not have a ton of individualization, but there are some options:

```{r, fig.width = 12, fig.height = 4, warning = FALSE}
data |> 
  filter(Id %in% c(204, 216, 218)) |> #choosing 3 ids
  gg_heatmap(fill.limits = c(0,NA)) #sets the upper limit to the max value#

data |> 
  filter(Id %in% c(204, 216, 218)) |> #choosing 3 ids
  gg_heatmap(unit = "5 mins") #changes the binning
```

Importantly, it has a handy doubleplot feature, either showing the `next`, or the `same` day repeated

```{r, fig.width = 16, fig.height = 4, warning = FALSE}
data |> 
  filter(Id %in% c(204, 216, 218))|> 
  gg_heatmap(doubleplot = "next", fill.limits = c(0, NA))
```

Finally, we can use heatmaps to produce an Actigram-like plot.

```{r, fig.width = 16, fig.height = 4, warning = FALSE}
data |> 
  filter(Id %in% c(204, 216, 218))|> 
  gg_heatmap(Variable.colname = MEDI >= 50, 
             doubleplot = "next", 
             fill.limits = c(0, NA), 
             fill.remove = TRUE, fill.title = ">50lx MEDI") + 
  scale_fill_manual(values = c("TRUE" = "black", "FALSE" = "#00000000"))
```


# gg_doubleplot()

A more generalized implementation of doubleplots is realized with `gg_doubleplot()`: it repeats days within a plot, either horizontally or vertically. Doubleplots are generally useful to visualize patterns that center around midnight (horizontally) or that deviate from 24-hour rhythms (vertically). 

## Preparation

We will use a subset from the data used in `gg_day()` above: two Ids, aggregated to 15-minute intervals, and restricted to three days. Because the first day is only partly present for both Ids, we will use the `gap_handler()` function to fill in the implicitly missing data with NA. If we ignore this step, the doubleplot will be incorrect, as it will connect the last point of the first day (around midnight) with the first point of the second day (somewhen before noon), which is incorrect and also looks bad.

```{r}
data_subset <- 
  data_subset |> 
  gap_handler(full.days = TRUE)
```

## Horizontal doubleplot

The horizontal doubleplot is activated by default, if only one day is present within all provided groups, or it can be set explicitly by `type = "repeat"`. 

```{r, fig.height=10}
 data_subset |> 
   gg_doubleplot(aes_fill = Id, jco_color = TRUE, type = "repeat")
#identical:
# data_subset |> 
#  group_by(Date = date(Datetime), .add = TRUE) |>
#  gg_doubleplot(aes_fill = Id, jco_color = TRUE)
```

Each plot line thus is the same day, plotted twice.

## Vertical doubleplot

The vertical doubleplot is activated by default if any group has more than one day. It can be set explicitly by `type = "next"`. 

```{r,fig.height=10}
data_subset |> 
  gg_doubleplot(aes_fill = Id, jco_color = TRUE)
#identical:
# data_subset |> 
#  gg_doubleplot(aes_fill = Id, jco_color = TRUE, type = "next")
```
Note that the second day in each row is the first day of the next row. This allows to visualize non-24-hour rhythms, such as when Entrainment is lost due to pathologies or experimental conditions. Note that the x-axis labels change automatically depending on whether the doubleplot has type `"next"` or `"repeat"`.

In both cases (horizontally and vertically) it is easy to condense the plots to a single line per day, by ungrouping the data structure (makes only sense if the datetimes are identical):

```{r, fig.height=5}
data_subset |> 
  ungroup() |> 
  gg_doubleplot(aes_fill = Id, jco_color = TRUE)
```

## Aggregated doubleplot

Independent of `gg_doubleplot()`, but great in concert with it is `aggregate_Date()`, which allows to aggregate groups of data to a single day each. This way, one can easily calculate the average day of a participant or a group of participants and then perform a doubleplot (by default with `type = "next"`).

Let us first group our data by whether participants were in the first or last two months of the experiment.

```{r}
data_two_groups <- data |> 
  mutate(
    Month = case_when(month(Datetime) %in% 8:9 ~ "Aug/Sep",
                      month(Datetime) %in% 10:11 ~ "Oct/Nov")
  ) |> 
  group_by(Month)
```

Now we can aggregate the data to a single day per group and make a doubleplot from it. With `aggregate_Date()` we condense a large dataset with 10-second intervals to a single day for two groups with a 15 minute interval. The day that is assigned by default is the median measurement day of the group.

```{r}
data_two_groups |> 
  aggregate_Date(unit = "15 mins") |> 
  gg_doubleplot(aes_fill = Month, jco_color = TRUE) +
  guides(fill = "none")
```

We can further improve this plot. First, we overwrite the default behavior by setting a specific (arbitrary) date for both groups. By ungrouping the data afterwards, we can plot the two groups in a single row, making the two times easily comparable. Setting `facetting = FALSE` gets rid of the strip label, that otherwise would show the (arbitrary) date. 

```{r}
data_two_groups |> 
  aggregate_Date(unit = "15 mins", 
                 date.handler = \(x) as_date("2023-09-15")
                 ) |> 
  ungroup() |> 
  gg_doubleplot(aes_fill = Month, jco_color = TRUE, facetting = FALSE)
```

We clearly see now, that daytime light exposure starts later and ends earlier, with lower daytime values overall. Conversely, nighttime light exposure is increased in the second half of the experiment (Oct/Nov). By using the `gg_doubleplot()` feature, the nighttime light pattern is clearly visible even accross midnight.

# Photoperiod

`LightLogR` contains a family of functions to calculate and work with photoperiods. This includes visualization, which is as easy as adding `gg_photoperiod()` to your choice of `gg_day()`, `gg_days()`, or `gg_doubleplot()`. Here is a minimal example:

```{r, fig.width = 9}
#specifying coordinates (latitude/longitude)
coordinates <- c(48.521637, 9.057645)

sample.data.environment |> 
  gg_days() |> 
  gg_photoperiod(coordinates) 
```

More information on dealing with photoperiods can be found in the article [Photoperiod](https://tscnlab.github.io/LightLogR/articles/photoperiod.html).

# gg_states()

`gg_states()` is another great add on plotting function that works similarly to `gg_photoperiod()` in that it is called on top of an existing plot. Let us start by filtering the larger dataset down to that participant.

## Preparation

```{r}
#filter the dataset
data_205 <-
  data |> filter(Id == "205") |> 
  aggregate_Datetime(unit = "5 mins", type = "floor")
```

Next we are importing sleep data for a participant (Id = 205), which is included in `LightLogR`:

```{r}
#the the path to the sleep data
path <- system.file("extdata", 
              package = "LightLogR")
file.sleep <- "205_sleepdiary_all_20230904.csv"
#import sleep/wake data
dataset.sleep <- 
  import_Statechanges(file.sleep, path, 
                      Datetime.format = "dmyHM",
                      State.colnames = c("sleep", "offset"),
                      State.encoding = c("sleep", "wake"),
                      Id.colname = record_id,
                      sep = ";",
                      dec = ",",
                      tz = "Europe/Berlin")
```

Next, we add the sleep/wake data to the dataset. While we are at it, we also add the Brown et al. 2022 recommendations for healthy light, which can be extracted from sleep/wake data.

```{r}
data_205 <-
  data_205 |>
  interval2state(dataset.sleep |> sc2interval()) |> #add sleep/wake-data
  interval2state(
    dataset.sleep |> sc2interval() |> sleep_int2Brown(), 
    State.colname = State.Brown) #add Brown et al. 2022 states
```

## With gg_days()

Adding the sleep-wake information to a base plot.

```{r, fig.width = 9}
data_205 |> gg_days() |> gg_states(State, aes_fill = State)
```

We can also only highlight sleep phases, either by setting `wake` instances to `NA`, or, by converting it to a logical column. Notice that a conditional fill is no longer necessary.

```{r, fig.width = 8}
data_205 |> 
  mutate(State = ifelse(State == "sleep", TRUE, FALSE)) |> 
  gg_days() |> 
  gg_states(State)
```

## With gg_day()

`gg_states()` automatically detects whether it is called from `gg_day()` or `gg_days()`, so it just works out of the box.

```{r, fig.height = 8}
data_205 |> 
  gg_day(geom = "line") |> 
  gg_states(State.Brown, aes_fill = State.Brown) +
  labs(fill = "Brown states")
```

We can even go a step beyond, and show at which times the participant complied with these recommendations. We will employ the helper function `Brown2reference()` that creates additional columns, including one that tests whether a light exposure level is within the required range (`Reference.check`). When we select this column for `gg_states()`, we can color by the `State.Brown` same as above, but get the selection of when this recommendation was actually met. 

```{r, fig.height = 8}
data_205 |> 
  Brown2reference() |>
  group_by(State.Brown, .add = TRUE) |> 
  gg_day(geom = "line") |> 
  gg_states(Reference.check, aes_fill = State.Brown) +
  labs(fill = "Brown states")
  
```

## With gg_doubleplot()
 
`gg_states()` should mostly work fine with `gg_doubleplot()` out of the gate

```{r, fig.height = 8}
data_205 |> 
  mutate(State = ifelse(State == "sleep", TRUE, FALSE)) |> 
  gg_doubleplot() |> 
  gg_states(State)
```

## Combination with other add-on functionality

State plotting can be combined with other add-on plotting functions, like other states, photoperiods, or even gaps. However it can quickly become confusing if there are overlapping states, like sleep and nighttime:

```{r, fig.height = 8}
data_205 |> 
  mutate(State = ifelse(State == "sleep", TRUE, FALSE)) |> 
  gg_doubleplot() |> 
  gg_states(State, aes_fil = State, aes_col =State, alpha = 0.1) |> 
  gg_photoperiod(c(48.5,9)) +
  labs(colour ="Sleep", fill = "Sleep")
```

In these cases it might be more useful to color the ribbon according to the state or the photoperiod.

### Emphasis ond photoperiod with sleep/wake

```{r, fig.height = 8}
data_205 |> 
  mutate(State = ifelse(State == "sleep", TRUE, FALSE)) |> 
  add_photoperiod(c(48.5,9)) |> 
  gg_doubleplot(aes_fill = photoperiod.state, group = consecutive_id(photoperiod.state)) |> 
  gg_states(State) +
  labs(fill = "Photoperiod")
```

### Emphasis on sleep/wake with photoperiod

```{r, fig.height = 8}
data_205 |> 
  gg_doubleplot(aes_fill = State, group = consecutive_id(State)) |> 
  gg_photoperiod(c(48.5,9)) +
  labs(fill = "Sleep")
```

# gg_gaps()

`gg_gaps()` visualizes gaps and optionally also irregular data. It is easy to use, but can be computationally expensive, if there are lots of irregular data and/or gaps. Calling it on good data does produce a plot

```{r}
data |> gg_gaps()
```

We can create a dataset with explicit and implicit gaps and irregular data, where all zero values are replaced with `NA`, observations above 1000 lx are missing, and the last day has a slightly delayed sequence.

```{r, fig.width = 8}
bad_dataset <-
data_205 |>
   mutate(Datetime = if_else(date(Datetime) == max(date(Datetime)),
                             Datetime, Datetime + 1), #creates irregular data for the last day
          MEDI = na_if(MEDI, 0) #creates explicit gaps
          ) |>
  filter(MEDI <1000) #creates implicit gaps

bad_dataset |> gg_gaps(MEDI)
```
By default, `gg_gaps()` only shows missing values. Setting `show.irregulars = TRUE` also adds the irregular data to the plot

```{r, fig.width = 8}
bad_dataset |> gg_gaps(MEDI, show.irregulars = TRUE)
```


# Interactivity

`gg_day()` and `gg_days()` have the inbuilt option to be displayed interactively. This is great for exploring the data. The `plotly` package is used for this. They have the `interactive` argument set to `FALSE` by default. Setting it to `TRUE` will create an interactive plot. 

```{r}
data_subset |>  
  gg_day(aes_col = Id, geom = "line",
         interactive = TRUE
         )
```

# Miscellaneous

## Non-Light properties

`LightLogR` is designed to work with light data, but it can also be used for other types of data. Simply define the y.axis argument in the plotting functions. In the following example, we will plot the activity data from the `data` dataset.

For comparison, the light data is added in the background with a lower alpha value.

```{r, fig.width=12}
data_subset2 |>
  gg_days(y.axis = PIM, y.axis.label = "Activity (PIM)") +
  geom_line(aes(y=MEDI), color = "red", alpha = 0.2)
```

## Scales

By default, `LightLogR` uses a so-called `symlog` scale for visualizations. This scale is a combination of a linear and a logarithmic scale, which is useful for light data, as it allows to visualize very low and very high values in the same plot, including 0 and negative values. As light data is regularly zero, and exact values between 0 and 1 lux are usually not as relevant for devices measuring up to 10^5 lx, this scale is more useful compared to a linear or logarithmic scaling.

The way `symlog` works is that there is a threshold up to which absolute values are kept linear, and beyond which they are transformed logarithmically. The default threshold is 1, which is a good choice for light data. However, this can be changed by setting the `threshold` argument in the plotting functions. See a full documentation of the `symlog` scale `symlog_trans()`.

Here is an example to show how the transformation is particularly useful for differences that cross zero. We will use the single-day doubleplot data from above.

```{r, fig.height = 6}
#dataset from above
data <- 
  data_two_groups |> 
    aggregate_Date(unit = "15 mins", 
                   date.handler = \(x) as_date("2023-09-15")
                   ) |> 
    ungroup()

#original doubleplot from above
original_db <- 
  data |> 
    gg_doubleplot(aes_fill = Month, jco_color = TRUE, facetting = FALSE) +
    guides(fill = "none")

#difference doubleplot, showing the average difference between the to phases
difference_db <- 
  data |> 
    select(Datetime, MEDI, Month) |> 
    pivot_wider(names_from = Month, values_from = MEDI) |> 
    gg_doubleplot(y.axis = `Oct/Nov`-`Aug/Sep`, facetting = FALSE,
                  y.axis.label = "difference (lx, MEDI)")

#plotting
original_db / difference_db

```

We can clearly see the difference in light exposure is crossing 0 several times. With the symlog scale no values are discarded, because they are outside the traditional range of a logarithmic scale. Should values below 1 lux be of interest, the parameters of the transformation can be adjusted.

```{r}
data |> 
  select(Datetime, MEDI, Month) |> 
  pivot_wider(names_from = Month, values_from = MEDI) |> 
  gg_doubleplot(y.axis = `Oct/Nov`-`Aug/Sep`, facetting = FALSE,
                y.axis.label = "difference (lx, MEDI)",
                y.scale = symlog_trans(thr = 0.001),
                y.axis.breaks = c(-10^(-2:5), 0, 10^(5:-2))
                )
```

